---
title: "Session 1 Notes"
author: "Tim Riffe"
date: "2022-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to R Markdown

Whatever I type freely like this, is just text. Nothing bad can happen to your document by just typing here. So take all the notes you want.

## Code chunks

Hot-key for making a code chunk: `Ctrl + Alt +i`

Inside the code chunk, we need to have valid `R` code. In order to build this document, whatever code is executed in the code blocks needs to successfully run. I argue that this forces the analysis to be reproducible, and this is a good artifact of working in markdown.
```{r}
# use the hash to indicate a note

# this is a little piece of R:
a <- c(1,2,3)
a
```

To execute the code live, you can:
1. select the code entirely and press `Ctrl + Enter`
2. put the cursor somewhere on the line and press `Ctrl + Enter`
3. Click the little green arrow in the upper right corner of the code chunk and it will run the whole chunk.

## code sharing
2 ways to share code for this workshop. 
1. I might paste small pieces in the Google Doc\
2. I can update a script in Markdown. To copy that text click on the file, and then on the button called `Raw`.

## Use earlier code

Code chunks are executed sequentially, and are therefore sequentially dependent (potentially), meaning that to use an object, it needs to have been created earlier, and earlier-created objects are available later
```{r}
a ^ 2
age <- 0:100
age + 2

age_sq <- age ^ 2
age_sq

age <- age + 1
age_sq
```

Objects retain their identities. Look at what we do using `age` above. We create `age_sq` using it, but when we then alter `age`, `age_sq` didn't change. That is, an object stores a result rather than a relationship that may have created it.

You can create (using `<-`) or destroy objects at will (`rm()`)
```{r}
rm(age)
```
## assignment `<-` or `=`


```{r}
b <- c(3,4,5)
b = c(2,3,4)
a == b
b <- c(1,3,4)
a == b

lm(a ~ b, singular.ok = FALSE)
?lm
```
Copy-pasted from the `lm` help file examples; We don't necessarily need to understand each piece of the help file example, we just need to see what the objects look like that we need to feed to the function we're interested in (`lm`). 
```{r}
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
lm.D9 <- lm(weight ~ group)
lm.D90 <- lm(weight ~ group - 1) # omitting intercept
```

I showed how to find out more about a function than is well-explained in the help file example. I typed `"r lm offset"` in Google and it brought me to the most pertinent Stack Overflow question / answer.

## Package installation and loading

Only install a package once. You can usually just use `install.packages()` for this. Load its functions into your workspace using `library()`. `install.packages()` wants you to give the package name in quotes. `library()` does not care if there are quotes or not.

`R` was of course founded by statisticians for doing statistics, but it is in fact a general language, and it has many subcommunities with a common goal. Demographers have made many `R` packages too. Looking at the way they code methods is indeed one way which *might* build intuition on how a demographic method works.
```{r, message = FALSE, warning = FALSE}
# install.packages("tidyverse")
library(tidyverse)
?filter
```

So, let's install a bunch of packages for the sake of later days in this workshop.

```{r}
# install.packages("janitor")
# install.packages("lubridate")
# install.packages("countrycode")
# install.packages("scales")
# install.packages("colorspace")
# install.packages("readxl")
# install.packages("vroom")
# install.packages("DemoDecomp")
```

Trick for commenting out a bunch of lines at once: `Ctrl + Shift c`
Note, you can make a multi-line cursor by holding down `Alt` and dragging the cursor down. That's how I removed and then replaced the hash character.

## basic `R` objects


```{r}
dim(a)
length(a)
```

Notes on matrices
```{r}
A <- matrix(a)
dim(A)

# elementwise product
A * a

# a matrix product looks like this:
A %*% a

# with higher matrix dim, need to be careful with
# left-vs-right sides of matrix algebra formulas.
B <- cbind(A, c(3,4,5))
dim(B)

a %*% B

```

Notes on `data.frame`. Each column of a `data.frame` can be a different kind of data (numeric, integer, date, character, factor, logical). Each column must be the same length (rectangular). This is the kind of data object that we'll be using the most, or rather its close cousin the `tibble`. This sort of data is somehow easier to work with when doing so-called *tidy* operations.
```{r}
my_factor <- factor(c("a","b","c"))
D <- data.frame(a, 
                x = 0:2, 
                y = runif(3), 
                group = c("a","b","c"),
                my_factor)
D
```





