---
title: "Session 1 Notes"
author: "Tim Riffe"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

We will take notes in R markdown, which is the kind of document that we're in here. In general, you can write text as you like here. i.e. in your language, it doesn't matter. You definitely don't need to write what I write verbatim.


To create a code chunk in a markdown document, type `Ctrl Alt i` (Cmnd Option i on mac)

```{r}
1+1
```

## R objects

`runif()` is a function for generating random deviates from the uniform distribution. We know it's a function because of the round parentheses.

A function in `R` is sort of like a function in math: you give it inputs, and it gives you back some result. In other words it maps the input to the result.
This analogy will only be helpful if you have memories from a math class. If not no worries, it is just as it sounds: you give it something, it gives you something back.
$$f()$$

`<-` assigns the result of the function call (the evaluation of `runif()`) to the name on the left.

```{r}
my_first_object <- runif(10)

my_first_object * 2
```

## help files

To see the help file for a function, just type `?` followed by the function name.
```{r}
?runif
```

## artithmetic

`Ctrl Enter` executes the line where the cursor is (or the text that is selected)

```{r}
2*3
2+3
2-3
2/3
2^3
my_first_object

# elements here are not independent
cumsum(my_first_object)
cumprod(my_first_object)

# each element is independent with these ones:
exp(my_first_object)
log(my_first_object)
```

Here's my definition of a cumulative sum. It's important to note that the elements of a vector have index positions, i.e. they are ordered in the sense of the first element being first.
$$ 
cumsum_i = \sum _{i=1}^n x_i
$$

## table-like objects
`my_first_object` is a vector with 10 elements, i.e. of length 10.

we can create a tabular object using `data.frame()`, by simple listing out the columns, comma separated. The name on the left of the `=` is the column name, and on the right is what it consists on. These column definitions are comma-separated. We end up with a table-like object with 10 rows and 3 columns. Note, the columns can have different data types. We have integer, double, and character.

```{r}
a <- data.frame(x = 1:10,
           y = my_first_object,
           z = letters[1:10])
str(a)
head(a)
```

`str()` tells us metadata about the object: the type (`data.frame`), the dimensions (10x3), and then a listing of the individual columns shoing the data types and the first few values.

`head()` gives us a look at the first 6 rows by default.

## writing out data:

If you want to write out to a file, for example to grab values in Excel or whatever, then I suggest using a semicolon as the separator. Commas will come back to bite you one day.

If you have a table-like object, then you can use `write.table()` and use the `sep` argument to give the delimiter. 
```{r}
write.table(a, file = "a.csv", sep = ";")
```

## grabbing material from github:

One way to get my script during class is to tell me to push it to GitHub. Then navigate to the file you want to see, click `raw` then select, copy, and paste.

## installing packages

Execute this line just once. We don't need to re-install every time we build this markdown file. That would just waste time.

`install.packages()` looks for the package you name in the CRAN repositories.

```{r, eval = FALSE}
install.packages("tidyverse")
```

To load a package, use `library()`

```{r, message = FALSE}
library(tidyverse)
```

```{r}
KOR <- read_csv("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/KOR2014.csv", show_col_types = FALSE)

head(KOR)

KOR |> 
  pull(sex) |> 
  unique()
```

I wanted to check whether we just had data for females, or whether we had a full sex breakdown of the population. It turns out we males, females, and total.

You can make this pipe: `|>` using `Ctrl Shift m` 
You pipe might look like this: `%>%`
The pipe reads as "and then do this"

`pull()` selects (extracts) a column, which should be named.
`unique()` lists the unique values of a vector.

This whole expression reads as a single sentence. "Take KOR then pull off the sex column and then list its unique values."

## `filter()` and `select()`

`filter()` is for rows. 
`==` is a logical equals, as opposed to an assignment equals. That means we're asking.

```{r}
2 == 2/1
my_first_object > .5

```
```{r}
b = 2
# same as:
b <- 2
```

`select()` is for columns. Negative selections throw out (delete) a column.
```{r}
KOR |> 
  select(-year)
```

```{r}
KOR |> 
  filter(sex == "female",
         between(age, 12,55)) |> 
  select(age, exposure, deaths)
```

## Let's look at this data

To create a plot using ggplot (gg stands for grammar of graphics). It's grammatical because there's a systematic way to map variables to aesthetics (coordinates, but also things like size, width, shape, color, etc).

`ggplot()` starts the plot. You should include an aesthetic mapping inside it, using `mapping = aes()`. Other aesthetic attributes are set rather than mapped. 

Then you need to declare a `geom`etric rendering. 
```{r}
KOR |> 
  filter(sex != "total") |> 
  ggplot(mapping = aes(x = age, y = deaths, color = sex)) +
  geom_point(size = 3) +
  theme_minimal()
```

## objective: plot mortality rates

In order to create a new column for mortality rates, we use `mutate()`. `mutate()` is for creating or modifying columns. Then we pipe into `ggplot()` and make the plot as before, except now we have the rates `mx` in the y coordinates, and we use a line geometry instead of point geometry, and we log the y axis in order to see the pattern.

What we see here is quite consistent with the death count patterns that we saw before. But still some of the difference will be due to population structure differences.
```{r}
KOR |> 
  filter(sex != "total") |> 
  mutate(mx = deaths / exposure) |> 
  ggplot(mapping = aes(x = age, y = mx, color = sex)) +
  geom_line() +
  scale_y_log10() +
  theme_minimal()
```




















