---
title: "Session 1 Notes"
author: "Tim Riffe"
date: "2022-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to R Markdown

Whatever I type freely like this, is just text. Nothing bad can happen to your document by just typing here. So take all the notes you want.

## Code chunks

Hot-key for making a code chunk: `Ctrl + Alt +i`

Inside the code chunk, we need to have valid `R` code. In order to build this document, whatever code is executed in the code blocks needs to successfully run. I argue that this forces the analysis to be reproducible, and this is a good artifact of working in markdown.
```{r}
# use the hash to indicate a note

# this is a little piece of R:
a <- c(1,2,3)
a
```

To execute the code live, you can:
1. select the code entirely and press `Ctrl + Enter`
2. put the cursor somewhere on the line and press `Ctrl + Enter`
3. Click the little green arrow in the upper right corner of the code chunk and it will run the whole chunk.

## code sharing
2 ways to share code for this workshop. 
1. I might paste small pieces in the Google Doc\
2. I can update a script in Markdown. To copy that text click on the file, and then on the button called `Raw`.

## Use earlier code

Code chunks are executed sequentially, and are therefore sequentially dependent (potentially), meaning that to use an object, it needs to have been created earlier, and earlier-created objects are available later
```{r}
a ^ 2
age <- 0:100
age + 2

age_sq <- age ^ 2
age_sq

age <- age + 1
age_sq
```

Objects retain their identities. Look at what we do using `age` above. We create `age_sq` using it, but when we then alter `age`, `age_sq` didn't change. That is, an object stores a result rather than a relationship that may have created it.

You can create (using `<-`) or destroy objects at will (`rm()`)
```{r}
rm(age)
```
```{r}

```




