---
title: "Day 1 session notes"
author: "Tim Riffe"
date: "7/26/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

You can use R as a little basic calculator:

```{r}
# This is R!!!!!
1 + 1
```

The short cut to create a code chunk is:

press these three keys on your keyboard:
Ctrl + Alt + i

```{r}

```

That created a blank code chunk! So we can stick code in it that gets executed. Which is the only thing we're going to do in this course!


## Including Plots

You can also embed plots, for example:

```{r echo = FALSE}
plot(pressure)
```

This space here is just for notes. What you write here is your business.
Your note-taking space is for your own benefit, and you should be writing here ideally in your work language or native language.

```{r, eval = FALSE}
# install.packages("DemoDecomp")
install.packages("readr")
install.packages("readxl")
install.packages("lubridate")
install.packages("scales")
install.packages("colorspace")
install.packages("tidyverse")

```

Installing packages gets them into your R libraries on your computer, but it does not make them directly available to you in your R session. To load a package, use `library()` and type the package name inside it with no quotes.

```{r}
library(tidyverse)
```

Back to basics. Here are the operators, there are more!
```{r}
1 + 1
1 / 2
1 * 2
1 - 2
5^2
# matrix multiplication is different
# %*%
```

Create an object. Here we created a vector using `c()`. It has three elements, so we call it a vector of length 3. We use the `length()` function to ask how many elements it has.

```{r}
x <- c(2, 5, 7)
length(x)
```

How can we look at data? (`View()`), or we can query metadata about an object using `str()` or `glimpse()`.

```{r}
View(x)
str(x)   # metadata
glimpse(x)
```

Using arithmetic functions on vectors.

```{r}
2 + 5 + 7
sum(x)
cumsum(x)
mean(x)
sd(x)
var(x)
exp(x)
log(x)
```

How to learn how functions work and what they do?

```{r}
?sum
y <- c(2,5,7,NA)
sum(y, na.rm = FALSE)
```

Here are the examples from `?sum` (the help file) pasted directly into a new R chunk.
```{r}
## Pass a vector to sum, and it will add the elements together.
sum(1:5)

## Pass several numbers to sum, and it also adds the elements.
sum(1, 2, 3, 4, 5)

## In fact, you can pass vectors into several arguments, and everything gets added.
sum(1:2, 3:5)

## If there are missing values, the sum is unknown, i.e., also missing, ....
sum(1:5, NA)
## ... unless  we exclude missing values explicitly:
sum(1:5, NA, na.rm = TRUE)
```

Lesson is: getting a help file is as easy as typing `?` and the function name. You can also search in the help tab in R Studio.

```{r}
x
is.vector(x)
class(x)
is.integer(x)

x <- as.integer(x)
is.integer(x)
length(x)
dim(x)
```

Cleaning up, use `rm()` to remove objects from your *workspace*

```{r}
rm(y)
```

There are also character vectors (strings):
```{r}
b <- c("A","1","c","b")
d <- c(TRUE, TRUE, FALSE, FALSE)
```


What about tabular data: rectangular data, or data organized in rows and columns, like in a spreadsheet:
```{r}
A <- data.frame(b, d)
A
```

A `data.frame()` has a dimension! `dim()` tells us rows and then columns.

```{r}
dim(A)
nrow(A) # ncol()
```

You can add rows to a `data.frame` using `rbind()`, `bind_rows()`. Rules for adding rows are that you need the same number and structure of columns in both pieces of data.

```{r}
B <- data.frame(b = c("adth","Tim"), 
                d = c(FALSE, FALSE))
# This is identical. No need to create intermediate objects.
# becuase what we're doing is small
# b2 <- c("adth", "Tim")
# d2 <- c(FALSE, FALSE)
# data.frame(b = b2, d = d2)

D <- rbind(A, B)
D <- bind_rows(A, B)
```


You can add a column to a `data.frame`:

```{r}
# old fashioned way:
D$z <- runif(6)

# Delete the column:
D$z <- NULL

# the tidy way:
z <- runif(6)
E <- bind_cols(D,z = z)

# tidy column deletion
select(E, -z)
```















