---
title: "Session 1 Notes"
author: "Tim Riffe"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

We will take notes in R markdown, which is the kind of document that we're in here. In general, you can write text as you like here. i.e. in your language, it doesn't matter. You definitely don't need to write what I write verbatim.


To create a code chunk in a markdown document, type `Ctrl Alt i` (Cmnd Option i on mac)

```{r}
1+1
```

## R objects

`runif()` is a function for generating random deviates from the uniform distribution. We know it's a function because of the round parentheses.

A function in `R` is sort of like a function in math: you give it inputs, and it gives you back some result. In other words it maps the input to the result.
This analogy will only be helpful if you have memories from a math class. If not no worries, it is just as it sounds: you give it something, it gives you something back.
$$f()$$

`<-` assigns the result of the function call (the evaluation of `runif()`) to the name on the left.

```{r}
my_first_object <- runif(10)

my_first_object * 2
```

## help files

To see the help file for a function, just type `?` followed by the function name.
```{r}
?runif
```

## artithmetic

`Ctrl Enter` executes the line where the cursor is (or the text that is selected)

```{r}
2*3
2+3
2-3
2/3
2^3
my_first_object

# elements here are not independent
cumsum(my_first_object)
cumprod(my_first_object)

# each element is independent with these ones:
exp(my_first_object)
log(my_first_object)
```

Here's my definition of a cumulative sum. It's important to note that the elements of a vector have index positions, i.e. they are ordered in the sense of the first element being first.
$$ 
cumsum_i = \sum _{i=1}^n x_i
$$

## table-like objects
`my_first_object` is a vector with 10 elements, i.e. of length 10.

we can create a tabular object using `data.frame()`, by simple listing out the columns, comma separated. The name on the left of the `=` is the column name, and on the right is what it consists on. These column definitions are comma-separated. We end up with a table-like object with 10 rows and 3 columns. Note, the columns can have different data types. We have integer, double, and character.

```{r}
a <- data.frame(x = 1:10,
           y = my_first_object,
           z = letters[1:10])
str(a)
head(a)
```

`str()` tells us metadata about the object: the type (`data.frame`), the dimensions (10x3), and then a listing of the individual columns shoing the data types and the first few values.

`head()` gives us a look at the first 6 rows by default.

## writing out data:

If you want to write out to a file, for example to grab values in Excel or whatever, then I suggest using a semicolon as the separator. Commas will come back to bite you one day.

If you have a table-like object, then you can use `write.table()` and use the `sep` argument to give the delimiter. 
```{r}
write.table(a, file = "a.csv", sep = ";")
```

## installing packages

```{r}

```













