---
title: "Tuesday session notes"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

Today we'll get an introduction to the tidyverse. Which includes reading in data, and reshaping it, recoding it, and merging data together. 

# Read in the births data

Tidy data (our objective) is defined as a tabular arrangement of data, where columns are strictly variables and rows consist in single observations.

```{r}
# packages we'll need
# install.packages("here")
library(tidyverse)
library(readxl)
library(here)
Wide <- read_excel(path = here("Data", "demo_fasec.xlsx"),
           range = "A10:H158")
glimpse(Wide)
```

`pivot_longer()` collects a range of columns and stacks them. `names_to` is where the previous columns names get collected in a new column. `values_to` is where the cell values are collected as a single column.
```{r}
Long <-
   pivot_longer(data = Wide,
                cols = 3:8,
                names_to = "TIME",
                values_to = "Births")
Long
```

The column range can be specified by name too, or also using various kinds of conditional selection. In the second example it chooses all columns where the data type is `double`.

```{r}
# select using column name range
 pivot_longer(data = Wide,
                cols = `2011`:`2016`,
                names_to = "TIME",
                values_to = "Births")

 pivot_longer(data = Wide,
                cols = where(is.double),
                names_to = "TIME",
                values_to = "Births")
```

Select and rename columns to whatever standard we want. When we assign to `Long` having started with `Long`, it overwrites the old one.
```{r}
Long <- 
  select(.data = Long,
       Country = `GEO/TIME`,
       Age = AGE,
       Year = TIME,
       Births)
glimpse(Long)
```

Now let's redo the above three steps making use of piping. `%>%` Ctrl + Shift + m

```{r}
Long <-
# step 1, read it in
  read_excel(
    path = here("Data", "demo_fasec.xlsx"),
    range = "A10:H158") %>% 
  
# step 2, stack the years
  pivot_longer(
    cols = `2011`:`2016`,
    names_to = "Year",
    values_to = "Births"
  ) %>% 
  
# step 3 select and rename columns the way we want
  select(Country = `GEO/TIME`,
         Year,
         Age = AGE,
         Births)
```









