---
title: "Session 2 Notes"
author: "Tim Riffe"
date: "2024-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

To use tidyverse functions, we need to load the package
```{r, message=FALSE}
library(tidyverse)
```

## Load HMD dataset

We got this data url by navigating to the file on github and copying the url by right-clicking on the `raw` button.

`read_csv()` is not the same as `read.csv()`

When we read this in, a bunch of metadata prints to the console, which we can toggle off by adding the `show_col_types = FALSE` argument.

```{r}
hmd <- read_csv("https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/hmd.csv.gz", 
                show_col_types = FALSE)
```

Take a look. Here are some helpful functions for getting to know an unfamiliar dataset:

```{r}
head(hmd) # look at first rows
str(hmd)
glimpse(hmd)

hmd$sex |> unique()
hmd$country |> unique()
```

## Visualize a subset

```{r}
hmd |> 
  filter(country == "France",
         year == 1945) |> 
  ggplot(mapping = aes(x = age, y = mx, linetype = sex)) +
  geom_line() +
  scale_y_log10()
```

## Visualize the ENTIRE dataset


```{r}
hmd |> 
  mutate(mx = if_else(mx == 0, NA, mx),
         half_century = year - year %% 50) |> 
  ggplot(aes(x = age, 
             y = mx, 
             group = interaction(sex, country, year))) +
  geom_line(alpha = .03) +
  scale_y_log10() +
  facet_grid(vars(half_century),vars(sex))
```


## explanation of modulo

```{r}
x <- 0:10
x %% 2
(1750:2000) - 1750:2000 %% 50
```

## Challenge nr 1

Take the following dataset in single ages and group the `value` column into 5-year age group bins.
You'll be using `%%`
and the following tidy functions: `mutate()`, `group_by()`, `summarize()`.
```{r}
A <- data.frame(age = 0:100, value = rpois(n = 101, lambda = 1000))

A |> 
  mutate(age = age - age %% 5) |> 
  group_by(age) |> 
  summarize(value = sum(value))
```

## functions

Anatomy of a function
1. it has a name (name on the left that gets assigned to)
2. assign the `function()` function to it.
3. list your arguments (parameters) in the parentheses of `function()`.
4. define the body between `{ body here }`
5. use your arguments to do something in the body- only use things that get passed in via arguments.
6. `return()` some sort of result.

```{r}
my_first_function <- function(arg1, arg2){
  step1 <- arg1 * arg2 ^ 2
  step2 <- sqrt(step1)
  
  return(step2)
}
my_first_function(arg1 = 10, arg2 = 5)

```

## demonstrat `cumprod()`

This is a vector-in vector-out function that calculates a cumulative product.
```{r}
a <- runif(10)
cumprod(a)

```

## Next example

$$ q_x = \frac{n \cdot M_x}{(1 - (n - a_x) \cdot M_x)}  $$

```{r}
mx_to_qx <- function(mx, ax, n = 1){
  num   <- (n * mx) 
  denom <- (1 - (n - ax) * mx)
  qx    <- num / denom
  n     <- length(qx)
  qx[n] <- 1
  return(qx)
}
```

Here, we split the formula into numerator and denominator steps, just for the sake of legibility. We can make objects inside the function, but the only thnig that gets out is what gets returned (ideally explicitly).

Example usage:
```{r}
hmd |> 
  filter(country == "Australia", 
         year == 2000,
         sex == "f") |> 
  mutate(qx = mx_to_qx(mx = mx,
                       ax = ax))
```

## survivorship

$$ \ell_x = \prod_0^{x-1} (1 - q_t)$$
where 
$$ \ell_0 = 1 $$
```{r}
qx_to_lx <- function(qx, radix = 1){
  
  # the essential bit
  lx <- c(1, cumprod(1 - qx))
  
  # ensure the same length
  lx <- lx[-length(lx)] 
  
  # scale it up, potentially
  lx <- lx * radix
  return(lx)
}
```

## the lifetable deaths distribution

$$ d_x = \ell_x \cdot q_x$$


```{r}
qxlx_to_dx <- function(qx, lx){
  dx <- lx * qx
  return(dx)
}
```


Here we use one of our own functions inside a custom function. This is called modularity.
```{r}
qx_to_dx <- function(qx, radix = 1){
  lx <- qx_to_lx(qx = qx, radix = radix)
  dx <- qx * lx
  return(dx)
}
```

## derive lifetable person years lived

$$ {}_nL_x = \ell_x\cdot n - {}_nd_x\cdot (n - {}_na_x)$$

```{r}
lxdx_to_Lx <- function(lx,dx,ax,n=1){
  Lx <- (lx * n) - (dx * (n - ax))
  return(Lx)
}

mx_to_Lx <- function(mx, ax, n = 1, radix = 1){
  qx <- mx_to_qx(mx = mx, ax = x, n = n)
  lx <- qx_to_lx(qx = qx, radix = radix)
  dx <- qxlx_to_dx(lx = lx, qx = qx)
  Lx <- lxdx_to_Lx(lx = lx, dx = dx, ax = ax, n = n)
  return(Lx)
}

```

## Total years lived above age x

$$ T_x = \sum _{t=x}^\omega {}_nL_t$$
```{r}
a <- 1:10
a
rev(a) # reverse a vector
cumsum(a) # cumulative sum

# rLx <- rev(Lx)
# rTx <- cumsum(rLx)
# Tx <- rev(rTx)

# rev(cumsum(rev(Lx)))
Lx_to_Tx <- function(Lx){
  Tx <- 
    Lx |> 
    rev() |> 
    cumsum() |> 
    rev()
    return(Tx)
}

```

## Remaining life expectancy at age x

$$ e_x = \frac{T_x}{\ell_x} $$

```{r}
Txlx_to_ex <- function(Tx, lx){
  ex <- Tx / lx
  return(ex)
}
```

## apply these functions to calculate a lifetable

```{r}
hmd |> 
  filter(country == "Australia", 
         year == 2000,
         sex == "f") |> 
  mutate(qx = mx_to_qx(mx = mx, ax = ax),
         lx = qx_to_lx(qx = qx, radix = 1),
         dx = qxlx_to_dx(qx = qx, lx = lx),
         Lx = lxdx_to_Lx(lx = lx, dx = dx, ax = ax, n = 1),
         Tx = Lx_to_Tx(Lx = Lx),
         ex = Txlx_to_ex(Tx = Tx, lx = lx)
         )
```

# Big challenge
1. Calculate lifetables for EVERY subset in the HMD data
2. make a scatterplot of life expectancy $e_0$ at birth against $q_0$
`geom_point()`

```{r}
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_functions_2024.R")
hmd |> 
  group_by(country, sex, year) |> 
  mutate(qx = mx_to_qx(mx = mx, ax = ax),
         lx = qx_to_lx(qx = qx, radix = 1),
         dx = qxlx_to_dx(qx = qx, lx = lx),
         Lx = lxdx_to_Lx(lx = lx, dx = dx, ax = ax, n = 1),
         Tx = Lx_to_Tx(Lx = Lx),
         ex = Txlx_to_ex(Tx = Tx, lx = lx)) |> 
  ungroup() |> 
  filter(age == 0) |> 
  ggplot(mapping = aes(x = ex, y = qx, color = sex)) +
  geom_point(alpha = .1) +
  xlim(20,90)
```




Or we could use a nice wrapper function that does the whole lifetable!
```{r}
hmd |> 
  group_by(country, sex, year) |> 
  group_modify(~calc_LT_tidy())
```











































