---
title: "Session 3 notes"
author: "Tim Riffe"
date: "2024-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Unobserved heterogeneity

Load some functions we need
```{r, message = FALSE}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_functions_2024.R")
```

Gompertz mortality
$$ \mu(x) = \alpha \cdot e^{\beta x}$$
Make a little Gompertz mx function:
```{r}
gomp_mx <- function(alpha, beta, age){
  mx <- alpha * exp(beta * age)
  return(mx)
}

# a cheap life expectancy at birth function:
mx_to_e0 <- function(mx){
  ax <- rep(.5, length(mx))
  e0 <- 
    calc_LT(mx = mx, 
          ax = ax, 
          n = 1, 
          radix = 1) |> 
    slice(1) |> 
    pull(ex)
  
  return(e0)
}

```

Invent two mortality schedules:

```{r}
alpha1 <- .0001
alpha2 <- .0004
beta   <- .075
age    <- 0:100

mx1    <- gomp_mx(alpha1, beta, age)
mx2    <- gomp_mx(alpha2, beta, age)

mx_to_e0(mx1)
mx_to_e0(mx2)
```

Simulate what would happen to the total population


```{r}
mx <- data.frame(group = rep(c("a","b"), 
                             each = 101),
           age = c(age, age),
           mx = c(mx1, mx2)) 
```

```{r}
mxt <-
  mx |> 
  # invent / impute ax and n needed by lifetable function
  mutate(ax = rep(.5, n()),
         n = rep(1, n())) |> 
  
  # create lifetables for each subset
  group_by(group) |> 
  group_modify(~calc_LT_tidy(data = .x)) |> 
  
  # Lots of unnecessary columns, just pick what we want
  select(group, age, mx, ax, Lx) |> 
  
  # assuming an initial compositon.
  mutate(Px = Lx * .5) |> 
  
  # next, calculate prevalence of each group
  # in each age
  ungroup() |> 
  group_by(age) |> 
  mutate(prevx = Lx / sum(Lx)) |> 
  
  # (groups still active) calculate weighted average mortality
  summarize(mx = sum(mx * prevx)) |> 
  # (summarize removes one of your groups always,
  # in this case we only had one, so we lose it)
  
  mutate(group = "t", .before = 1)
```

Join the data and plot the phenomenon
```{r}
mx |> 
  bind_rows(mxt) |> 
  ggplot(mapping = aes(x = age, y = mx, color = group)) +
  geom_line() +
  scale_y_log10()
```

## Calculate POADR

POADR = Prospective Old-Age Dependency Ratio
(Sanderson and Scherbov, multiple papers)

1. Load WPP 2022 data
There are 2 datasets on the workshop github site
1.1 `pop1dt.csv.gz`
1.2 `mx1dt.csv.gz`

Load these into `R` using `read_csv()`. Call the first object `pop` and the second object `mx`.

```{r, message = FALSE}
pop <- read_csv('https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/pop1dt.csv.gz')
mx <- read_csv('https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/mx1dt.csv.gz')

head(pop)
```



2. Reshape each of these objects to tidy format. As-is, the data columns are `mxM`, `mxF`, and `mxB`. To do this, you'll use the `pivot_longer()` function, which I have mentioned but not demonstrated. Here is an example of its use:

```{r}
A <- tibble(country = rep("a",6),
            year = rep(2020,6),
            mx_a = runif(6),
            mx_b = runif(6))
A # now we want to stack mx:
A |> pivot_longer(-c(country, year), 
                  names_to = "subgroup", 
                  values_to = "mx")
```



2. Let's add the `ex` column of the lifetable to the `mx` object.
Throw out the `popM`, `popF` columns of `pop` and remove the `MxM` and `MxF` columns of `mx`.

```{r}
pop <-
  pop |> 
  select(-popM, -popF)

mx <-
  mx |> 
  select(-mxM, -mxF)
# join the datasets
wpp <- left_join(pop, mx, by = c("country_code",
                                 "name",
                                 "year","age"))
```

# Generate the `ex` column of the lifetable
I suggest doing this in a `mutate()` call, going one
column at a time, but you could also apply our lifetable wrapper function as earlier today, up to you. When done, remove the columns we no longer need.

```{r}
ex_example <- 
wpp |> 
  #group_by(country_code, year) |> 
  filter(country_code == 900, year == 2000) |> 
  mutate(ax = if_else(age == 0, .1, .5),
         qx = mx_to_qx(mx = mxB, ax = ax, n = 1),
         lx = qx_to_lx(qx = qx, radix = 1),
         dx = qxlx_to_dx(qx = qx, lx = lx),
         Lx = lxdx_to_Lx(lx = lx, dx = dx, ax = ax, n = 1),
         Tx = Lx_to_Tx(Lx = Lx),
         ex = Txlx_to_ex(Tx = Tx, lx = lx)) |> 
  select(country_code, name, year, age, pop, ex) #|> 
  # ggplot(aes(x = age, y = ex)) +
  # geom_line() +
  # geom_hline(yintercept = 15)
```

## make a POADR threshold age generator function

```{r}

ex <- ex_example |> pull(ex)
age <- ex_example |> pull(age)
pop  <- ex_example |> pull(pop)

lower_bound <- 15

POADR_linear <- function(age, 
                         ex, 
                         pop, 
                         upper_threshold = 15, 
                         lower_bound = 15){
# approx(x = ex, y = age, xout = 15)
P15 <- approxfun(x = ex, y = age)(upper_threshold)

split_age <- floor(P15)
frac      <- P15 - split_age

denom <- sum(pop[between(age, lower_bound, (split_age - 1))]) +
         pop[age == split_age] * frac

numerator <- sum(pop[age > split_age]) +
  pop[age == split_age] * (1 - frac)

POADR <- numerator / denom
# splinefun(x = ex, y = age)(15)
return(POADR)
}

POADR_linear(age, ex, pop, upper_threshold = 15, lower_bound = 15)

```

Challenge:
Make an OADR function. `pop`, `age`, `lower_bound` and `upper_bound` should be its arguments.

$$ OADR = \frac{\sum_{x=65}^{\omega}P_x}{\sum_{x=15}^{64}P_x}$$
```{r}

OADR_calc <- function(age, 
                 pop, 
                 upper_bound = 65, 
                 lower_bound = 15){

  numerator <- pop[age >= upper_bound] |> 
    sum()
  denominator <- pop[between(age, 
                             lower_bound, 
                             (upper_bound - 1))] |> sum()
 DR <- numerator / denominator
 return(DR)
}

```

Final challenge of today:
Make a scatterplot (`geom_point()`) of OADR vs POADR.
That means scale up the calculation of `ex`, then recycle the same groups to calculate the dependency measures inside a `summarize()` call.

```{r}
oadr <- wpp |> 
  group_by(country_code, name, year) |> 
  
  # remove anomalous cases
  mutate(keep = !any(is.na(mxB)) &
           mxB[1] < .15) |> 
  ungroup() |> 
  filter(keep) |> 
  
  # generate ex (this group_by() possibly extraneous)
  group_by(country_code, name, year) |> 
  mutate(ax = if_else(age == 0, .1, .5),
         qx = mx_to_qx(mx = mxB, ax = ax, n = 1),
         lx = qx_to_lx(qx = qx, radix = 1),
         dx = qxlx_to_dx(qx = qx, lx = lx),
         Lx = lxdx_to_Lx(lx = lx, dx = dx, ax = ax, n = 1),
         Tx = Lx_to_Tx(Lx = Lx),
         ex = Txlx_to_ex(Tx = Tx, lx = lx)) |> 
  select(country_code, name, year, age, pop, ex) |> 
  
  # calculate indicators (this group_by() possibly extraneous)
  group_by(country_code, name, year) |> 
  reframe(POADR = POADR_linear(age = age,
                                pop = pop,
                                ex = ex,
                                lower_bound = 15,
                                upper_threshold = 15),
            OADR = OADR_calc(age = age, 
                             pop = pop, 
                             lower_bound = 15,
                             # there was a mistake here before,
                             # I accidentally put 15 for both bounds!
                             upper_bound = 65))
```

plot it!
```{r}
oadr |> 
  ggplot(aes(x = OADR, y= POADR, color =year))+
  geom_point(alpha=.05) +
  theme_minimal() +
  geom_abline(slope=1,intercept=0,color="red")
```














