---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2022-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

```{r}

hello <- function(your_name){
  paste0("Hello, ", 
         your_name, 
         "! Have a nice day!")
}
paste0("a","b")
hello(your_name = c("Tim","Sugyeong","Gowoon"))
```

## a numerical example

```{r}
my_mathy_function <- function(x){
  35 + 3 * x^2 + x*sin(x^2)
}
x <- seq(0,100,by=.01)
my_mathy_function(x)

tibble(x, 
       y = my_mathy_function(x)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line()
```

### function of n arguments

Note

1. arguments `x`, `y`, and `z` should all be the same length, or just one element.
2. you can use other functions inside your functions. They also can be YOUR functions.
3. Whatever is evaluated last in the function gets returned. However, it's better to explicitly return using `return()`.
4. You can and should annotate code inside of functions.
```{r}
a_bigger_function <- function(x, y, z){
  x <- my_mathy_function(x = x)
  x * y^z
}

a_bigger_function <- function(x, y, z){
  # apply earlier-written function
  x   <- my_mathy_function(x = x)
  out <- x * y^z
  return(out)
}

# a_bigger_function(x = 4, y = 2, z = 13:76)
```

# The lifetable

```{r}
library(tidyverse)
mx <- read_csv("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/LT_inputs.csv",
               show_col_types = FALSE)
mx$Country %>% unique()
mx$Age %>% unique()
mx$Year %>% unique()
```

```{r}
mx %>% 
  filter(Year == 2000,
         Country == "Algeria") %>% 
  ggplot(mapping = aes(x = Age, y = nMx, color = Sex)) +
  geom_line() +
  scale_y_log10()
```
Pull off a few vectors that can be our toy data for purposes of designing our lifetable functions.
```{r}

# pull of nMx
nmx <- mx %>% 
  filter(Country == "Algeria",
         Year == 2000,
         Sex == "f") %>% 
  pull(nMx)

nax <- mx %>% 
  filter(Country == "Algeria",
         Year == 2000,
         Sex == "f") %>% 
  pull(nAx)

n <- c(1,4,rep(5, 17))
```

Ready to convert to `nqx`. You can find this very formula in the Preston book, or the HMD methods protocol.
$$ {}_nq_x = \frac{n \cdot {}_nM_x}{1 + (n - {}_nA_x) \cdot {}_nM_x} $$
Let's make a function whose arguments are `nmx`, `nax`, and `n`.
```{r}
nmx_to_nqx <- function(nMx, nAx, n){
  nqx <- (n * nMx) / ( 1 + (n - nAx) * nMx)
  
  # these are called "kludge"s
  nqx[nqx > 1] <- 1
  nqx[nqx < 0] <- 0
  
  return(nqx)
}
nqx <- nmx_to_nqx(nMx = nmx,
           nAx = nax,
           n = n)
```

## `lx`

$$ l_x = R \cdot \prod_{y=0}^x (1 - {}_nq_y)   $$
```{r}
nqx_to_lx <- function(nqx, radix){
  # cumulative product of the conditional
  # survival probabilities in each age.
  lx <- cumprod(1 - nqx)
  lx <- c(1, lx)
  lx <- lx[-length(lx)]
  lx <- radix * lx
  return(lx)
}

tibble(age = age,
       lx = nqx_to_lx(nqx, radix = 1)) %>% 
  ggplot(aes(x = age, y = lx)) +
  geom_line() +
  ylim(0,1)

```

### Exercise (4 min) lifetable deaths

$$ {}_nd_x = {}_nq_x \cdot l_x $$

```{r}
nqxlx_to_ndx <- function( nqx, lx){
  nqx * lx
}
```

### Exercise (5 min) Person-years lived between $x$ and $x+n$

$$ {}_nL_x = n \cdot l_x - (n - {}_nA_x)\cdot{}_nd_x  $$

```{r}
get_Lx <- function(lx,nAx,ndx,n){
  n * lx - (n - nAx) * ndx
}
```


### Exercise (5 min) Person-years lived over age $x$

$$ T_x = \sum _{y=x}^\omega {}_nL_y $$

```{r}
get_Tx <- function(nLx){
  nLx %>% rev() %>% cumsum() %>% rev()
}
```

$$ e_x = \frac{T_x}{l_x} $$
```{r}
get_ex <- function(Tx, lx){
  Tx / lx
}
```

# Reboot on 2nd day!

```{r}

```


