---
title: "session 4 notes"
author: "Tim Riffe"
date: "2024-07-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Stationary population

```{r, message = FALSE}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_functions_2024.R")

mx <- read_csv('https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/mx1dt.csv.gz')
```

Note, lx is proportional to what the stationary population structure would be if the same number of births entered each year and mortality never changed over time.

```{r}
world <-
  mx |> 
  filter(country_code == 900,
         year == 2019) |> 
  select(-mxB) |> 
  rename(m = mxM, f = mxF) |> 
  pivot_longer(c(m,f),
               names_to = "sex",
               values_to = "mx") |> 
  arrange(sex, age)
```

Now we have a clean subset that can be used for later calculations
First, derive Lx
```{r}
SRB <- 1.05
(PF <- 1 / (1 + SRB))
stationary <-
  world |> 
  group_by(sex) |> 
  mutate(ax = if_else(age == 0, .1, .5),
         n = rep(1,n())) |> 
  group_modify(~calc_LT_tidy(data = .x)) |> 
  select(sex, age, Lx) |> 
  # Cx is structure, implying a shared total radix of 1
  mutate(Cx = Lx * if_else(sex == "f", PF, 1 - PF)) |> 
  ungroup() |> 
  mutate(Cx = 100 * Cx / sum(Cx))
```

Plot it!

```{r}
stationary |> 
  mutate(Cx = Cx * if_else(sex == "m", -1, 1)) |> 
  ggplot(aes(x = age, y = Cx, fill = sex)) +
  geom_col(width = 1, alpha = .6) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(-.75, .75, by = .25),
                     labels = c(".75",".50",".25","",".25",".50",".75")) +
  geom_hline(yintercept = seq(-.75, .75, by = .25), 
             color = "white",
             linewidth = .5) +
  scale_x_continuous(breaks = seq(0,100,by=20)) +
  geom_vline(xintercept = seq(0,100,by=20), 
             color = "white", 
             linewidth = .5) +
  coord_flip() +
  scale_fill_manual(values = c(m = "#5c9c6c", f = "#965c9c"))
```

What about 5-year age groups?

```{r}
0:10 - 0:10 %% 5
```


```{r}
stationary |> 
  mutate(age = age - age %% 5) |> 
  group_by(sex, age) |> 
  summarize(Cx = sum(Cx), .groups = "drop") |> 
  mutate(Cx = Cx * if_else(sex == "m", -1, 1) / 5) |> 
  
  ggplot(aes(x = age, y = Cx, fill = sex)) +
  geom_col(width = 5, alpha = .6) +
  theme_minimal() +
  geom_hline(yintercept = seq(-.75, .75, by = .25), 
             color = "white",
             linewidth = .5) +
  scale_x_continuous(breaks = seq(0,100,by=20)) +
  geom_vline(xintercept = seq(0,100,by=20), 
             color = "white", 
             linewidth = .5) +
  coord_flip() +
  scale_fill_manual(values = c(m = "#5c9c6c", f = "#965c9c")) +
  guides(fill = "none")
```

# some properties of stationary populations
1. Population size $P$ is the number of births (= the radix size) times life expectancy

$$ P = e(0) \cdot \ell(0) = b \cdot e(0)$$

2. The crude birth rate $b$ is the crude death rate $m$ is the inverse of life expectancy.

$$ b =  \frac{1}{e(0)} = m$$

3. The age-standardized death rate is the same as the crude death rate (and the previously listed things too)

$$ b = m =\sum _0^\omega C(x) \mu(x) = ASDR = \frac{1}{e(0)}$$

4. The age structure (years lived) is equal to the time-to-death structure (years left)

$$ C(x) = \frac{\ell(x)}{e(0)} = \frac{\ell^\star(y)}{e(0)} = C^\star (y)$$
(where y stands for remaining years of life). That means that if we invert the pyramid (per the Wednesday structure presentation), we'd end up with the EXACT same pyramid again. 

5. In a stationary population with heterogeneity (like strata), we get the same equalities within strata groups.

6. In a stationary population with transient heterogeneity, analogous findings hold.

Why do we care about these rather hypothetical relationships? 
1. They are permanent
2. In some circumstances stationarity is not distant, like for fast-moving demographic parameters held fixed. For example the prevalence of a condition with rather stable patterns of onset and recovery will turn out to be very close in empirical data to the one you would get in a stationary population.

## stable populations

$$ C(x,r) = \frac{C(x)\cdot e^{-rx}}{\sum C(x)\cdot e^{-rx}} $$

```{r}
r <- seq(-.03,.03, by = .005)
growth_rates <- tibble(r = r)

stable <- 
  stationary |> 
  cross_join(growth_rates) |> 
  arrange(r, sex, age) |> 
  mutate(Cxr = Cx * exp(-r * age)) |> 
  group_by(r) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr))
```

Plot it!

```{r}
library(colorspace)
stable |> 
  mutate(Cxr = Cxr * if_else(sex == "m", -1, 1)) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cxr, 
                       color = r, 
                       group = interaction(sex,r))) +
  geom_line(linewidth = 1) +
  coord_flip() +
  scale_color_binned_diverging("Blue-Red") +
  theme_minimal()
```

```{r}
growth_rates <- tibble(r = c(-.005,-.01,-.02), P =55000000)
horizon <- tibble(horizon = 0:200)
P <- cross_join(growth_rates, horizon) |> 
  mutate(Pt = P * exp(r * horizon))

P |> 
  ggplot(aes(x = horizon, y = Pt, color = r, group = r)) +
  geom_line() +
  geom_hline(yintercept=10000000) +
  theme_minimal() 
```

# cohorts and periods in stable population models.

Note, in a stationary time population over time, a cohort's age structure over time is equal to the period age structure. In a stable (growing or shrinking) population, the $e^{-rT}$ discounting (inflation) that we do is all for purposes of making a period snapshot: Each period is proportional, but of a different size, or constant successive ratio over time. However, each _cohort_ selected from a stable time series is proportional to the stationary (lifetable survivorship) that was used to generate it. This means that cohorts can acheive stable and predictable age structures over time ... ooooo. That might be important when designing institutions and policies. Pay-as-you-go systems will probably perish in fast-shrinking populations, whereas a system build in the cohort perspective would be stable over time.

I think this is worth demonstrating just so that I can make this point.

```{r}
r <- -.03
stable <- 
  stationary |> 
  filter(sex == "f") |> 
  mutate(Cxr = Cx * exp(-r*age)) |> 
  cross_join(horizon) |> 
  rename(year = horizon) |> 
  arrange(year, age) |> 
  mutate(Cxr = Cxr * exp(r * year))
```


The time series we start with
```{r}
stable |> 
  ggplot(aes(x = year, y = age, z = Cxr)) +
  geom_contour_filled() +
  labs("a fast-shrinking population")
```


First, note each year has the same structure

```{r}
stable |> 
  group_by(year) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr)) |> 
  ggplot(aes(x = age, y = Cxr, linetype = as.factor(year), color = year)) +
  geom_line() +
  guides(linetype = "none", color = "none") +
  labs(title = "each year has the same structure")
```

Verify there are in fact different lines above
```{r}
stable |> 
  group_by(year) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr)) |> 
  ggplot(aes(x = year, y = age, z = Cxr)) +
  geom_contour_filled()
```


If you extract the cohorts, they are all proportional to stationarity
```{r}
stable |> 
  mutate(cohort = year - age) |> 
  filter(between(cohort, 0, 90),
         cohort %% 10 == 0) |> 
  group_by(cohort) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr)) |> 
  ggplot(aes(x = age, 
             y = Cxr, 
             color = cohort, linetype = as.factor(cohort))) +
  geom_line() +
  guides(color = "none", linetype = "none")
```

Periods rescaled, to see period-changes in structure (same as overlapping lines for cohort structure 2 plots ago)- these are stationary: proportional to the original Lx. i.e. with no over-sized age groups.
```{r}
stable |> 
  mutate(cohort = year - age) |> 
  filter(between(cohort, 0, 90)) |> 
  group_by(cohort) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr)) |> 
  ggplot(aes(x = year, y = age, z = Cxr)) +
  geom_contour_filled() +
  coord_equal()
```


# Fertility Measures (from Day 2)

```{r}
KOR <- read_csv("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/KOR2014.csv",
                show_col_types = FALSE)
E <- KOR |> 
  filter(sex == "female") |> 
  select(age, exposure)

B <- KOR |> 
  filter(sex == "total") |> 
  select(age, births)
```

Calculate age-specific fertility rates
```{r}
fert <-
  left_join(E, B, by = "age") |> 
  filter(births > 0) |> 
  mutate(fx = births / exposure)
```

Plot the rates
```{r}
fert |> 
  ggplot(aes(x = age, y = fx)) +
  geom_line()
```

Calculate the mean age at childbearing using fx as weights. Remember, weight together age + 1/2 rather than just age. Otherwise the result would be as if people only gave birth on their own birthdays? We use half, for half an interval, assuming, as demographers often do, that events happen _midway_ into time intervals.

```{r}
fert |> 
  summarize(MAB = sum(age * fx) / sum(fx) + .5,
            MAB2 = sum(age * B) / sum(B) + .5)

fert |> 
  ggplot(aes(x = age, y = fx)) +
  geom_step()
```

I drew a picture here. I described the fertility distribution as a smoother. Namely, it spans a range of ages and is itself smooth. And so, even if it is interacting with relatively rough population structure, the outgoing cohort sizes will having diminishing roughness. i.e. fluctuations will tend to attenuate all else take equal.

```{r}
fert |> 
  summarize(TFR = sum(fx))
```

## Mortality projection primer

```{r}
mx |> 
  filter(name == "Australia",
         between(year, 1950, 2019)) |> 
  ggplot(aes(x = year, y = age, z = log(mxF))) +
  #geom_tile() +
  geom_contour_filled() +
 # scale_fill_continuous_sequential("PinkYl", trans = "log") +
  coord_equal()
```

```{r}
mx |> 
  filter(name == "Australia",
         between(year, 1950, 2019),
         age %% 10 == 0) |> 
  ggplot(aes(x = year, y = mxF, color = age, group = age)) +
  geom_line() +
  scale_y_log10()
```

# example of edges that can make this challenging

Pick out a year with sharp mortality changes over age in younger ages. This is where mortality projections tend to fail the most. Notice
```{r}
mx |> 
  filter(name == "Australia",
         year == 1980) |> 
  ggplot(aes(x = age, y = mxM)) +
  geom_line() +
  scale_y_log10()
``` 












