---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2022-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

```{r}

hello <- function(your_name){
  paste0("Hello, ", 
         your_name, 
         "! Have a nice day!")
}
paste0("a","b")
hello(your_name = c("Tim","Sugyeong","Gowoon"))
```

## a numerical example

```{r}
my_mathy_function <- function(x){
  35 + 3 * x^2 + x*sin(x^2)
}
x <- seq(0,100,by=.01)
my_mathy_function(x)

tibble(x, 
       y = my_mathy_function(x)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line()
```

### function of n arguments

Note

1. arguments `x`, `y`, and `z` should all be the same length, or just one element.
2. you can use other functions inside your functions. They also can be YOUR functions.
3. Whatever is evaluated last in the function gets returned. However, it's better to explicitly return using `return()`.
4. You can and should annotate code inside of functions.
```{r}
a_bigger_function <- function(x, y, z){
  x <- my_mathy_function(x = x)
  x * y^z
}

a_bigger_function <- function(x, y, z){
  # apply earlier-written function
  x   <- my_mathy_function(x = x)
  out <- x * y^z
  return(out)
}

# a_bigger_function(x = 4, y = 2, z = 13:76)
```

# The lifetable

```{r}
library(tidyverse)
# mx <- read_csv("Data/LT_inputs.csv")
mx <- read_csv("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/LT_inputs.csv", show_col_types = FALSE)
mx$Country %>% unique()
mx$Age %>% unique()
mx$Year %>% unique()
```

```{r}
mx %>% 
  filter(Year == 2000,
         Country == "Algeria") %>% 
  ggplot(mapping = aes(x = Age, y = nMx, color = Sex)) +
  geom_line() +
  scale_y_log10()
```
Pull off a few vectors that can be our toy data for purposes of designing our lifetable functions.
```{r}

# pull of nMx
nmx <- mx %>% 
  filter(Country == "Algeria",
         Year == 2000,
         Sex == "f") %>% 
  pull(nMx)

nax <- mx %>% 
  filter(Country == "Algeria",
         Year == 2000,
         Sex == "f") %>% 
  pull(nAx)

n <- c(1,4,rep(5, 17))
```

Ready to convert to `nqx`. You can find this very formula in the Preston book, or the HMD methods protocol.
$$ {}_nq_x = \frac{n \cdot {}_nM_x}{1 + (n - {}_nA_x) \cdot {}_nM_x} $$
Let's make a function whose arguments are `nmx`, `nax`, and `n`.
```{r}
nmx_to_nqx <- function(nMx, nAx, n){
  nqx <- (n * nMx) / ( 1 + (n - nAx) * nMx)
  
  # these are called "kludge"s
  nqx[nqx > 1] <- 1
  nqx[nqx < 0] <- 0
  
  return(nqx)
}
nqx <- nmx_to_nqx(nMx = nmx,
           nAx = nax,
           n = n)
```

## `lx`

$$ l_x = R \cdot \prod_{y=0}^x (1 - {}_nq_y)   $$
```{r}
nqx_to_lx <- function(nqx, radix){
  # cumulative product of the conditional
  # survival probabilities in each age.
  lx <- cumprod(1 - nqx)
  lx <- c(1, lx)
  lx <- lx[-length(lx)]
  lx <- radix * lx
  return(lx)
}

tibble(age = age,
       lx = nqx_to_lx(nqx, radix = 1)) %>% 
  ggplot(aes(x = age, y = lx)) +
  geom_line() +
  ylim(0,1)

```

### Exercise (4 min) lifetable deaths

$$ {}_nd_x = {}_nq_x \cdot l_x $$

```{r}
nqxlx_to_ndx <- function( nqx, lx){
  nqx * lx
}
```

### Exercise (5 min) Person-years lived between $x$ and $x+n$

$$ {}_nL_x = n \cdot l_x - (n - {}_nA_x)\cdot{}_nd_x  $$

```{r}
get_Lx <- function(lx,nAx,ndx,n){
  n * lx - (n - nAx) * ndx
}
```


### Exercise (5 min) Person-years lived over age $x$

$$ T_x = \sum _{y=x}^\omega {}_nL_y $$

```{r}
get_Tx <- function(nLx){
  nLx %>% rev() %>% cumsum() %>% rev()
}
```

$$ e_x = \frac{T_x}{l_x} $$
```{r}
get_ex <- function(Tx, lx){
  Tx / lx
}
```

# Reboot on 2nd day!

```{r}
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/04_lifetable_functions.R")

library(tidyverse)
mx <- read_csv("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/LT_inputs.csv", show_col_types = FALSE)
```


Once we load our proto-library (`source("script.R")`) then we can freely use our functions without needing to re-copy them into this code. We do this because it's clean and simpler the hand-copying the new code into here. The difference between the functions we load and the ones above this point in the file are the names mostly. And in some cases the closeout details. Closeout details means "how we handle the highest age in the lifetable".

We show how our lifetable transformation functions can be used in a single call to `mutate()`, which makes it easy to do calculations in bulk.
```{r}
mx %>% 
  group_by(Country, Year, Sex) %>% 
  mutate(nqx = get_nqx(nMx = nMx,
                       nAx = nAx,
                       n = AgeInt),
         lx = get_lx(nqx = nqx,
                     radix = 100000),
         ndx = get_ndx(nqx = nqx,
                       lx = lx),
         nLx = get_nLx(lx = lx,
                 ndx = ndx,
                 nAx = nAx,
                 n = AgeInt),
         Tx = get_Tx(nLx = nLx),
         ex = get_ex(Tx = Tx,
                     lx = lx)) 
```

## Turn the `mutate()` part into a function

We have copied the `mutate()` part of the pipeline above into a function,
where we gave the incoming data a generic name `data`, and we assume the columns needed have the same names as what our lifetable minifunctions are anticipating, `nMx`, `aX`, and `AgeInt`, and we also made `radix` a top-level argument with a default value.
```{r}
my_lifetable <- function(Data, radix = 100000){
  out <-
   Data %>% 
     mutate(
     # convert death rates to probabilities
         nqx = get_nqx(nMx = nMx,
                       nAx = nAx,
                       n = AgeInt),
     # derive survival function
         lx = get_lx(nqx = nqx,
                     radix = radix),
     # derive lifetable deaths distribution
         ndx = get_ndx(nqx = nqx,
                       lx = lx),
     # lifetable exposure
         nLx = get_nLx(lx = lx,
                 ndx = ndx,
                 nAx = nAx,
                 n = AgeInt),
     # total exposure above exact age x
         Tx = get_Tx(nLx = nLx),
     # conditional average remaining lifetime
         ex = get_ex(Tx = Tx,
                     lx = lx)) 
  return(out)
}
```

How do we use this function. Note, we have "data frame in, data frame out".

Originally I assumed that `.data` was the name of the imaginary chunk of data being passed, but indeed it's `.x`, HT JinSeok Song for noticing.

```{r}
mx %>% 
  group_by(Country, Year, Sex) %>% 
  group_modify(~my_lifetable(Data = .x,
                             radix = 1)) %>% 
  ungroup()

```









