---
title: "Session 4 notes"
author: "Tim Riffe"
date: "7/29/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `ggplot2` basics

```{r}
# install.packages("gapminder")
library(gapminder)
library(tidyverse)

gapminder
```

The fundamentals, `ggplot()` is necessary but not sufficient to build the plot. here we declare, at minimum, the coordinate mapping. Executing this opens a blank plot area, which is however aware of the data dimensions.
```{r}
gapminder %>% 
  ggplot(mapping = aes(x = gdpPercap, y = lifeExp))
```

To actually *draw* something (sorry, map something) to the plot, we need to specify a geometric element, using `geom_*`
```{r}
gapminder %>% 
  ggplot(mapping = aes(x = gdpPercap, y = lifeExp)) +
  
  # this declare the geometric mapping
  geom_point()
```

If we instead specify `geom_line()` it renders the observations as a single path, because it doesn't know where each country begins and ends.
```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  #geom_line()
  geom_point() + 
  scale_x_log10() +
  scale_y_log10()
```



