---
title: "session 4 notes"
author: "Tim Riffe"
date: "2024-07-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Stationary population

```{r, message = FALSE}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_functions_2024.R")

mx <- read_csv('https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/mx1dt.csv.gz')
```

Note, lx is proportional to what the stationary population structure would be if the same number of births entered each year and mortality never changed over time.

```{r}
world <-
  mx |> 
  filter(country_code == 900,
         year == 2019) |> 
  select(-mxB) |> 
  rename(m = mxM, f = mxF) |> 
  pivot_longer(c(m,f),
               names_to = "sex",
               values_to = "mx") |> 
  arrange(sex, age)
```

Now we have a clean subset that can be used for later calculations
First, derive Lx
```{r}
SRB <- 1.05
(PF <- 1 / (1 + SRB))
stationary <-
  world |> 
  group_by(sex) |> 
  mutate(ax = if_else(age == 0, .1, .5),
         n = rep(1,n())) |> 
  group_modify(~calc_LT_tidy(data = .x)) |> 
  select(sex, age, Lx) |> 
  # Cx is structure, implying a shared total radix of 1
  mutate(Cx = Lx * if_else(sex == "f", PF, 1 - PF)) |> 
  ungroup() |> 
  mutate(Cx = 100 * Cx / sum(Cx))
```

Plot it!

```{r}
stationary |> 
  mutate(Cx = Cx * if_else(sex == "m", -1, 1)) |> 
  ggplot(aes(x = age, y = Cx, fill = sex)) +
  geom_col(width = 1, alpha = .6) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(-.75, .75, by = .25),
                     labels = c(".75",".50",".25","",".25",".50",".75")) +
  geom_hline(yintercept = seq(-.75, .75, by = .25), 
             color = "white",
             linewidth = .5) +
  scale_x_continuous(breaks = seq(0,100,by=20)) +
  geom_vline(xintercept = seq(0,100,by=20), 
             color = "white", 
             linewidth = .5) +
  coord_flip() +
  scale_fill_manual(values = c(m = "#5c9c6c", f = "#965c9c"))
```

What about 5-year age groups?

```{r}
0:10 - 0:10 %% 5
```


```{r}
stationary |> 
  mutate(age = age - age %% 5) |> 
  group_by(sex, age) |> 
  summarize(Cx = sum(Cx), .groups = "drop") |> 
  mutate(Cx = Cx * if_else(sex == "m", -1, 1) / 5) |> 
  
  ggplot(aes(x = age, y = Cx, fill = sex)) +
  geom_col(width = 5, alpha = .6) +
  theme_minimal() +
  geom_hline(yintercept = seq(-.75, .75, by = .25), 
             color = "white",
             linewidth = .5) +
  scale_x_continuous(breaks = seq(0,100,by=20)) +
  geom_vline(xintercept = seq(0,100,by=20), 
             color = "white", 
             linewidth = .5) +
  coord_flip() +
  scale_fill_manual(values = c(m = "#5c9c6c", f = "#965c9c")) +
  guides(fill = "none")
```

# some properties of stationary populations
1. Population size $P$ is the number of births (= the radix size) times life expectancy

$$ P = e(0) \cdot \ell(0) = b \cdot e(0)$$

2. The crude birth rate $b$ is the crude death rate $m$ is the inverse of life expectancy.

$$ b =  \frac{1}{e(0)} = m$$

3. The age-standardized death rate is the same as the crude death rate (and the previously listed things too)

$$ b = m =\sum _0^\omega C(x) \mu(x) = ASDR = \frac{1}{e(0)}$$

4. The age structure (years lived) is equal to the time-to-death structure (years left)

$$ C(x) = \frac{\ell(x)}{e(0)} = \frac{\ell^\star(y)}{e(0)} = C^\star (y)$$
(where y stands for remaining years of life). That means that if we invert the pyramid (per the Wednesday structure presentation), we'd end up with the EXACT same pyramid again. 

5. In a stationary population with heterogeneity (like strata), we get the same equalities within strata groups.

6. In a stationary population with transient heterogeneity, analogous findings hold.

Why do we care about these rather hypothetical relationships? 
1. They are permanent
2. In some circumstances stationarity is not distant, like for fast-moving demographic parameters held fixed. For example the prevalence of a condition with rather stable patterns of onset and recovery will turn out to be very close in empirical data to the one you would get in a stationary population.

## stable populations

$$ C(x,r) = \frac{C(x)\cdot e^{-rx}}{\sum C(x)\cdot e^{-rx}} $$

```{r}
r <- seq(-.03,.03, by = .005)
growth_rates <- tibble(r = r)

stable <- 
  stationary |> 
  cross_join(growth_rates) |> 
  arrange(r, sex, age) |> 
  mutate(Cxr = Cx * exp(-r * age)) |> 
  group_by(r) |> 
  mutate(Cxr = 100 * Cxr / sum(Cxr))
```

Plot it!

```{r}
library(colorspace)
stable |> 
  mutate(Cxr = Cxr * if_else(sex == "m", -1, 1)) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cxr, 
                       color = r, 
                       group = interaction(sex,r))) +
  geom_line(linewidth = 1) +
  coord_flip() +
  scale_color_binned_diverging("Blue-Red") +
  theme_minimal()
```

```{r}
growth_rates <- tibble(r = c(-.005,-.01,-.02), P =55000000)
horizon <- tibble(horizon = 0:200)
P <- cross_join(growth_rates, horizon) |> 
  mutate(Pt = P * exp(r * horizon))

P |> 
  ggplot(aes(x = horizon, y = Pt, color = r, group = r)) +
  geom_line() +
  geom_hline(yintercept=10000000) +
  theme_minimal() 
```


















##  






