---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2023-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# simple growth

# Geometric
$$ N(t) = N(0)\cdot(1+r)^t $$

This sort of growth could be used for intercensal population estimates in the case where no vital registration data were available between the censuses (meaning no cohort component estimates are available). In this case, connect the cohorts. This might require fiddling with the census estimates in advance to (i) adjust them, or (ii) align them so that ages cleanly correspond with birth cohorts. Then it doesn't matter if the census is not on a clean Jan 1, July 1 or Dec 31 reference date, because you can interpolate as you like to the reference dates that you need.
```{r}
N0   <- 1000
r    <- .05
time <- 100
Nt   <- N0 * (1 + r)^0:100
```

# Exponential
$$ N(t) = N(0)\cdot e^{rt} $$
```{r}
N0 * exp(r*time)
```

# stationary age structure

```{r, message = FALSE}
library(tidyverse)

hmd <- read_csv("https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/hmd.csv.gz",
                show_col_types = FALSE)
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_utils.R")
```

Let's pick out a subset and make a pyramid for it?

```{r}
SRB <- 1.05
PF  <- 1 / (1 + 1.05)


stationary <-
  hmd |> 
  filter(country == "Spain", year == 1970) |> 
  group_by(sex) |> 
  mutate(n = 1,
         qx = calc_qx(mx, ax, n),
         lx = calc_lx(qx),
         dx = calc_dx(lx, qx),
         Lx = calc_Lx(lx, dx, ax, n)) |> 
  ungroup() |> 
  select(sex, age, Lx) |> 
  # make the sex ratio conform to the sex ratio at birth
  mutate(Cx = if_else(sex == "f", Lx * PF, Lx * (1 - PF)),
  # recale to sum to 100%
         Cx = 100 * Cx / sum(Cx))
```

Why do we need `coord_flip()` For standard pyramids? It's because most pyramids have irregular profiles and are not monotonic. That means the order of y does not follow the order of x
```{r}
stationary |> 
  mutate(Cx = if_else(sex == "m", -Cx, Cx)) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       fill = sex)) +
  geom_col(width = 1) +
  # why coord_flip()? See note above
  coord_flip() +
  theme_minimal() +
  
  # all for the sake of grid lines in intuitive units / areas
  scale_y_continuous(labels = c(".6",".4",".2","",".2",".4",".6"), 
                     breaks = seq(-.6,.6,by=.2)) +
  scale_x_continuous(breaks = seq(0,100,by=25))+
  scale_fill_manual(values = c(m = "#abd98f", f = "#d9af8f")) +
  geom_hline(yintercept  = seq(-.6,.6,by=.2), 
             col = "white", 
             linewidth = .5) +
  geom_vline(xintercept = seq(25, 75, by = 25), 
             col = "white", 
             linewidth = .5) +
  
  # turn of fill legend
  guides(fill = "none") 
```

# Extra step for 5-year age group pyramids
Note, if you're plotting 5-year age group data, then we need to divide values by 5. This is because area is what matters in the plot. So if we don't divide by five, then the visual area will add up to 500\%. And then we can't overlay grouped and single-age data. This is a good technique for when comparing sources or versions. But note, if differences are merely subtle, then a different visualization would be better.

```{r}
stationary |> 
  # replace each single age with its lower bound 
  # by subtracting the remainder (modulo %%)
  mutate(age = age - age %% 5) |> 
  # group the data for aggregation
  group_by(sex, age) |> 
  summarize(Cx = sum(Cx)) |> 
  # adjust sex ratios and divide out bar widths
  # if age groups are irregular, then male a variable
  # for age_int, or something like that
  mutate(Cx = if_else(sex == "m", -Cx, Cx),
         Cx = Cx / 5) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       fill = sex)) +
  geom_col(width = 5) +
  coord_flip() +
  theme_minimal() +
  
  # all for the sake of grid lines in intuitive units / areas
  scale_y_continuous(labels = c(".6",".4",".2","",".2",".4",".6"), 
                     breaks = seq(-.6,.6,by=.2)) +
  scale_x_continuous(breaks = seq(0,100,by=25))+
  scale_fill_manual(values = c(m = "#abd98f", f = "#d9af8f")) +
  geom_hline(yintercept  = seq(-.6,.6,by=.2), 
             col = "white", 
             linewidth = .5) +
  geom_vline(xintercept = seq(25, 75, by = 25), 
             col = "white", 
             linewidth = .5) +
  
  # turn of fill legend
  guides(fill = "none") 
```


# stable age structure

The numerator and denominator have the same info, we just divide out the sum so that $C(x,r)$ adds to one and can be treated as a probability.

$$ C(x,r) = \frac{\ell(x)\cdot e^{-rx}}{\int_{x=a}^\omega\ell(a)\cdot e^{-ra} \mathrm{d}a}$$

```{r}
r <- -.005
stationary |> 
  # take care of discounting and scaling
  mutate(Cx = Cx * exp(-r * (age + .5)),
         Cx = 100 * Cx / sum(Cx),
         Cx = if_else(sex == "m", -Cx, Cx)) |> 
  # rest is the same
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       fill = sex)) +
  geom_col(width = 1) +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = c(".6",".4",".2","",".2",".4",".6"), 
                     breaks = seq(-.6,.6,by=.2)) +
  scale_x_continuous(breaks = seq(0,100,by=25))+
  scale_fill_manual(values = c(m = "#abd98f", f = "#d9af8f")) +
  guides(fill = "none") +
  geom_hline(yintercept = seq(-.6,.6,by=.2), 
             col = "white", 
             linewidth = .5) +
  geom_vline(xintercept = seq(25,75,by=25), 
             col = "white", 
             linewidth = .5)
```

`cross_join()` can join datasets that don't even have any column in common. That means the dataset expands. Here we make more copies of `stationary`, one for each values of `r`.
```{r}
r_table <- tibble(r = seq(-.03,.03,by=.005))

stable <-
  stationary |> 
  cross_join(r_table) |> 
  # now do the discounting, no need to group for this,
  # each age row has all the info it needs and we have no
  # inter-row dependencies to worry about
  mutate(Cx = Cx * exp(-r * (age + .5))) |> 
  # for comparison, rescale within each value of r 
  # so that we can compare.
  group_by(r) |> 
  mutate(Cx = 100 * Cx / sum(Cx)) |> 
  ungroup()
```

Visualize the stable structures. For this, we introduce the `colorspace` package. This gives a set of nice palettes to choose from. They all derive from science and are widely used both for their perceptual properties and because they look nice too. Use `hcl_palettes(plot = TRUE)` to view available palettes. You can load more palettes from other packages. Just be sure to pick one by name. We know we need a diverging palette for this case, since r spans positives and negatives. Only use divergent palettes when there's a clear transition point (in this case sign).

```{r}

# install.packages("colorspace")
library(colorspace)
stable |> 
  # arrange(sex,r,age) |> 
  mutate(Cx = if_else(sex == "m", -Cx, Cx)) |> 
  ggplot(mapping = aes(y = Cx, 
                       x = age, 
                       color = r, 
                       groups = interaction(sex, r))) +
  geom_line(alpha=.8) +
  coord_flip() +
  
  # scale color continuous diverging would have 
  # given us a contiuous color interp; binned picks
  # out just the needed color points
  scale_color_binned_diverging("Berlin")
```


# demonstrate period - cohort mismatch in stable populations

Here we will generate a time series of a stable population. We can do this
in two ways:
1.    By scaling a series of stationary age profiles by a time series of r, where each birth cohort is scaled by a single r.
2.    Going by periods, we scale using r in both the age and period directions! This is harder to wrap one's mind around, because going up age we use -r and going forward in time we use +r.

In the end these two approaches should give back the same Lexis surface of population structure.

```{r}
calendar_time <- tibble(cohort = 0:200)
r             <- -.01

stable_lexis <-
  stationary |> 
  group_by(age) |> 
  summarize(Cx = sum(Cx), .groups = "drop") |> 
  # same trick, except each cohort gets 1 r
  cross_join(calendar_time) |> 
  # order the rows
  arrange(cohort,age) |> 
  # perform the discount on entire cohorts,
  # back out period coordinate 
  mutate(Cx = Cx * exp(r * cohort),
         period = cohort + age)
```

Visualize the Lexis surface. `grom_tile()` draws every point. For a smooth filled contour, see below.
```{r}
stable_lexis |> 
  ggplot(mapping = aes(x = period, 
                       y = age, 
                       fill = Cx)) +
  geom_tile() +
  scale_fill_continuous_sequential("Sunset") +
  coord_equal() +
  theme_minimal()
```

Now we demonstrate that each vertical cross section will be proportional and stable, whereas each cohort cross section will be proportional but stationary.

```{r}
stable_lexis |> 
  filter(cohort %in% c(0,50,100)) |> 
  group_by(cohort) |> 
  mutate(Cx2 = Cx / sum(Cx)) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx2, 
                       color = cohort, 
                       group = cohort, 
                       linetype = as.factor(cohort))) +
  geom_line() +
  scale_color_binned_sequential("Emrld")
```

Each period cross section is stable and identical
```{r}
stable_lexis |> 
  filter(period %in% c(120,140,160)) |> 
  group_by(period) |> 
  mutate(Cx = Cx / sum(Cx)) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       color = period, 
                       group = period, 
                       linetype = as.factor(period))) +
  geom_line() +
  scale_color_binned_sequential("Emrld")
```

# Motivation for Lee Carter

Look how regular this is. Can you imagine bending a sheet of paper to conform to it? That's the idea, think simple.
```{r}

hmd |> 
  filter(country == "Australia",
         sex == "f") |> 
  ggplot(mapping = aes(x = year, 
                       y = age, 
                       z = log(mx))) +
  geom_contour_filled() 
  # scale_fill_continuous_sequential("OrYel",trans = "log")
```

First, log selected age patterns

```{r}
hmd |> 
  filter(country == "Australia",
         sex == "f",
         year %% 10 == 0) |> 
  ggplot(mapping = aes(x = age, 
                       y = mx, 
                       color = year, 
                       group = year)) +
  scale_y_log10() +
  scale_color_continuous_sequential("OrYel") +
  geom_line()
```

Do age groups develop linearly over time (log)
```{r}
hmd |> 
  filter(country == "Australia",
         sex == "f",
         age %% 10 == 0) |> 
  ggplot(mapping = aes(x = year, 
                       y = mx, 
                       color = age, 
                       group = age)) +
  scale_y_log10() +
  scale_color_continuous_sequential("OrYel") +
  geom_line()
```

end



