---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2022-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

```{r}

hello <- function(your_name){
  paste0("Hello, ", 
         your_name, 
         "! Have a nice day!")
}
paste0("a","b")
hello(your_name = c("Tim","Sugyeong","Gowoon"))
```

## a numerical example

```{r}
my_mathy_function <- function(x){
  35 + 3 * x^2 + x*sin(x^2)
}
x <- seq(0,100,by=.01)
my_mathy_function(x)

tibble(x, 
       y = my_mathy_function(x)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line()
```

### function of n arguments

Note

1. arguments `x`, `y`, and `z` should all be the same length, or just one element.
2. you can use other functions inside your functions. They also can be YOUR functions.
3. Whatever is evaluated last in the function gets returned. However, it's better to explicitly return using `return()`.
4. You can and should annotate code inside of functions.
```{r}
a_bigger_function <- function(x, y, z){
  x <- my_mathy_function(x = x)
  x * y^z
}

a_bigger_function <- function(x, y, z){
  # apply earlier-written function
  x   <- my_mathy_function(x = x)
  out <- x * y^z
  return(out)
}

# a_bigger_function(x = 4, y = 2, z = 13:76)
```


```{r}

```



