---
title: "session 4 notes"
author: "Tim Riffe"
date: "2025-06-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#

```{r}

a1 <- .00008
a2 <- .0003
b <-.07
age <- 30:100
gompmx <- function(a,b,x){
  a * exp(b * x)
}
mx_to_lx_cheap <- function(mx){
  n <- length(mx)
  c(1,exp(-cumsum(mx[-n])))
}
mx1 <- gompmx(a1,b,age)
mx2 <- gompmx(a2,b,age)
lx1 <- mx_to_lx_cheap(mx1)
lx2 <- mx_to_lx_cheap(mx2)


lx_tidy <- tibble(age, lx1, lx2)
prev30 <- c(rep(.01,70), rep(.1,7), rep(.2,7), rep(.3,7), rep(.4,70))

cohorts <- seq_along(prev30) + 1900
cohorts_tidy <- tibble(cohort = cohorts, prev_high = prev30)
library(tidyverse)

lx_data <- cross_join(cohorts_tidy,lx_tidy) 
prev_data <- 
lx_data |> 
  mutate(prev1 = prev_high * lx1,
         prev2 = (1 - prev_high) * lx2,
         year = cohort + age) |> 
  select(year, age, prev1, prev2) |> 
  group_by(year) |> 
  mutate(n = n()) |> 
  ungroup() |> 
  filter(n == 71) |> 
  select(-n)

prev_data |> 
  ggplot(aes(x = year, y = age, fill = prev1)) +
  geom_tile() +
  coord_equal()


```

The above code was shared via a script to be sourced from grithub.

# load lifetable functions
```{r}
# gets the lifetable function
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/refs/heads/master/lifetable_functions2025.R")

# run the prevalence simulation
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/refs/heads/master/04_simulation.R")

```

```{r}
mx_national <-  
  # collect edu-specific mx in table
  tibble(age = age, mx1 = mx1, mx2 = mx2) |> 
  # join to each year of edu prevalence
  right_join(prev_data, by = join_by(age)) |> 
  arrange(year, age) |> 
  # calculate national average mx
  mutate(mx = mx1 * prev1 + mx2 * prev2) |> 
  select(year, age, mx) |> 
  mutate(ax = .5)
```

# Exercise:
1. Calculate lifetables for each year (use `calc_LT()` inside of `reframe()`)
2. filter down to age 30
3. Make a graph of remaining life expectancy at age 30.

```{r}
mx_national |> 
  group_by(year) |> 
  reframe(calc_LT(mx = mx, ax = ax, age = age)) |> 
  filter(age == 30) |> 
  ggplot(mapping = aes(x = year, y = ex + 30)) +
  geom_line() +
  theme_minimal()
```

# load wpp mortality rates

```{r}
mxwpp <- read_csv("https://github.com/timriffe/KOSTAT_Workshop1/raw/refs/heads/master/Data/mx1dt.csv.gz")

```

# append country names

```{r}
library(countrycode)
mxwpp <-
  mxwpp |> 
  mutate(iso3 = countrycode(sourcevar = country_code,
                            origin = "un",
                            destination = "iso3c")) |> 
  filter(!is.na(iso3))
```

# reshape data to long format

```{r}
mxwpp <- 
  mxwpp |> 
  pivot_longer(starts_with("mx"), 
               names_to = "sex",
               values_to = "mx") |> 
  mutate(sex = case_when(sex == "mxM" ~ "male",
                         sex == "mxF" ~ "female",
                         TRUE ~ "total")) |> 
  arrange(iso3, year, sex, age)
```

# figure out infant a0

```{r}
mxwpp <-
  mxwpp |> 
  filter(sex != "total") |> 
  mutate(ax = case_when(
    sex == "male" & age == 0 & between(mx, 0, 0.02300) ~ 0.14929 - 1.99545 * mx,
    sex == "male" & age == 0 & between(mx, 0.02300001, 0.08307) ~ 0.02832 + 3.26021 * mx,
    sex == "male" & age == 0 & between(mx, 0.08307001, 1) ~ 0.29915,
    sex == "female" & age == 0 & between(mx, 0, 0.01724) ~ 0.14903 - 2.05527 * mx,
    sex == "female" & age == 0 & between(mx, 0.01724001, 0.06891) ~ 0.04667 + 3.88089 * mx,
    sex == "female" & age == 0 & between(mx, 0.06891001,1) ~ 0.31411,
    age == max(age) ~ 1 / mx, # constant mortality for closeout (the mean survival time)
    TRUE ~ .5                 # if subject to constant mortality is the inverse of the rate
  ))
```

# Exercise: calculate all lifetables.






