---
title: "Session 5 Notes"
author: "Tim Riffe"
date: "2023-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Projection

Pick out some data
```{r}
library(tidyverse)
library(colorspace)
hmd_url <- "https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/hmd.csv.gz"
hmd <- read_csv(hmd_url, 
                show_col_types = FALSE)

AUS <- hmd |> 
  filter(country == "Australia",
         sex == "f") |> 
  mutate(lmx = log(mx))
```

## Take a quick look:

```{r}
AUS |> 
  ggplot(mapping = aes(x = year, 
                       y = age, 
                       z = lmx)) +
  geom_contour_filled() +
  coord_equal()
```

# derive alpha

To calculate $\alpha(x)$ just take the mean over time of the log mortality rates in each age. We can do this using `group_by()` and `summarize()`. The line we get, we will use as-is. Notice it's already very smooth, but we risk amplifying any deviations (little bumps and divets) if we project this over time. This gets more serious with $\beta(x)$

```{r}
alpha <- 
  AUS |> 
  group_by(age) |> 
  summarize(alphax = mean(lmx))

# take a look
alpha |> 
  ggplot(mapping = aes(x = age, y = alphax)) +
  geom_line() +
  theme_minimal()
```

# get centered log mortality as matrix
```{r}
cent_l <-
  AUS |> 
  left_join(alpha,by="age") |> 
  mutate(lmx_centered = lmx - alphax) |> 
  select(age, year, lmx_centered) 

cent_l |> 
  ggplot(mapping = aes(x = year, y = age, z = lmx_centered)) +
  geom_contour_filled()+
  coord_equal()

cent_l |> 
  filter(age - age %% 10 == 0) |> 
  ggplot(mapping = aes(x = year, y = lmx_centered, color = age, group = age)) +
  geom_line() +
  scale_color_binned_sequential("Inferno")

```
Turn it into a matrix
```{r}
cenlM <-
  cent_l |> 
  pivot_wider(names_from = year,
              values_from = lmx_centered) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

# perform SVD

```{r}
svd_lM <- svd(cent_lM)
str(svd_lM)

U <- svd_lM$u
D <- svd_lM$d
V <- svd_lM$v

# can be recombined:
(U %*% diag(D) %*% t(V)) - cent_lM

```

# derive beta and kappa

```{r}
u      <- svd_lM$u[, 1]
v      <- svd_lM$v[, 1]
d      <- svd_lM$d[1]

Bx     <- u / sum(u)

# secular change
Kt     <- v * sum(u) * d
```

# put these back in tidy tables:

```{r}
beta <- 
  tibble(betax = Bx,
       age = 0:110)

kappa <- 
  tibble(kappat = Kt,
         year = 1921:2020)
```

# combine to predict:

```{r}
cross_join(beta, kappa) |> 
  left_join(alpha, by = "age") |> 
  mutate(lmx_lc = alphax + betax * kappat) |> 
  left_join(AUS, by = c("age","year")) 
```

# calculate drift and project kappa
```{r}
lK      <- length(Kt)
drift   <- (Kt[lK] - Kt[1]) / (lK - 1)
h       <- 1:100
Kt.fcst <- Kt[lK] + drift * h
```

# recombine to create projected mortality schedule

```{r}

```

