---
title: "Session 5 Notes"
author: "Tim Riffe"
date: "2025-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Get data

```{r, message = FALSE}
library(tidyverse)
hmd <- read_csv("https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/hmd.csv.gz")

source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_functions2025.R")

```

# Visualize surface

```{r}
ausf <- hmd |> 
  filter(country == "Australia",
         sex == "f") 

ausf |> 
  ggplot(aes(x = year, y = age, z = log(mx))) +
  geom_contour_filled() +
  coord_equal()
```

# Select year range to fit to

```{r}
ausf <-
  ausf |> 
  filter(year > 1975)
```


# derive alpha

```{r}
alpha <- 
  ausf|> 
  mutate(lmx = log(mx)) |> 
  group_by(age) |> 
  summarize(alphax = mean(lmx))
```

# Look at alpha and how it relates to our original mortality
```{r}
ausf |> 
  ggplot(aes(x = age, y = log(mx))) +
  geom_line(mapping = aes(group = year), alpha = .2) +
  geom_line(data = alpha, 
            mapping = aes(x = age, y = alphax),
            color = "red",
            linewidth = 2)
```

# make centered log mortality

```{r}
clmx <-
  ausf |> 
  mutate(lmx = log(mx)) |> 
  left_join(alpha, by = "age") |> 
  mutate(clmx = lmx - alphax)
```

Take a look at how clmx is different from lmx

```{r}
library(colorspace)
clmx |> 
  ggplot(aes(x = age, 
             y = clmx, 
             color = year, 
             group = year)) +
  geom_line() +
  scale_color_continuous_sequential("Inferno") #+
  # better look at how much we shrunk variation:
  #ylim(-12,0)
```

Make the matrix

```{r}
CLMX <-
  clmx |> 
  select(year, age, clmx) |> 
  pivot_wider(names_from = year,
              values_from = clmx) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

# perform SVD

```{r}
duv <- svd(CLMX)

str(duv)
d <- duv$d
U <- duv$u
V <- duv$v
# can be recombined:
# (U %*% diag(d) %*% t(V)) - CLMX
```

# derive beta and kappa

```{r}
#u     first column of U
u <- U[,1]
#v     first column of V
v <- V[,1]
#d     first element of d
d <- d[1]
#Bx    rescale u
betax <- u / sum(u)
# the average trend
kappat     <- v * sum(u) * d
```

# put beta and kappa in tidy tables:

```{r}

beta <- tibble(age = 0:110,
               betax = betax)
kappa <- tibble(year = 1976:2020,
                kappat = kappat)
```

Visualize the components

Beta:
```{r}
beta |> 
  ggplot(aes(x = age, y = betax)) +
  geom_line()
```
Kappa:
```{r}
kappa |> 
  ggplot(aes(x = year, y = kappat)) +
  geom_line() #+
  #geom_smooth(method = "lm")
```


# join and predict:

$$ \widehat{log(m(x))} = \alpha(x) + \beta(x)\cdot\kappa(t)$$

```{r}
mx_pred <- 
  kappa |> 
  cross_join(beta) |> 
  left_join(alpha, by = "age") |> 
  relocate(age, .after = year) |> 
  mutate(lmx_hat = alphax + betax * kappat) |> 
  left_join(clmx, by = c("year","age"))
```

# examine the residuals

```{r}
mx_pred |> 
  mutate(resid = lmx_hat - lmx,
         resid = case_when(resid > .3 ~ .3,
                           resid < -.3 ~ -.3,
                           TRUE ~ resid)) |> 
  ggplot(aes(x = year, y = age, fill = resid)) +
  geom_tile() +
  scale_fill_continuous_diverging("Blue-Red") +
  coord_equal()
```

# calculate drift and project kappa

```{r}
lk <- length(kappat)
drift <- (kappat[lk] - kappat[1]) /
  (lk - 1)

horizon <- 1:20
kappa_future <- kappat[lk] + (horizon * drift)

kappa_fut <-
  tibble(year = 2020 + horizon,
         kappat = kappa_future)
```

# Examine the joined kappa series

I see no anomaly here in the joining of kappa
```{r}
kappa |> 
  bind_rows(kappa_fut) |> 
  ggplot(aes(x=year, y = kappat)) +
  geom_line() +
  geom_point(data = kappa |> 
               filter(year == 2020), 
             size = 2,
             color = "red")
```


# recombine to create projected mortality schedule

```{r}
lc_pred <- 
  kappa |> 
  bind_rows(kappa_fut) |> 
  cross_join(beta) |> 
  left_join(alpha, by = "age") |> 
  mutate(lmx_hat = alphax + betax * kappat)
```

# check age patterns

```{r}
lc_pred |> 
  filter(year > 2020) |> 
  ggplot(aes(x = age, y = lmx_hat, color = year, group = year)) +
  geom_line() +
  scale_color_continuous_sequential("Inferno")
```

# project life expectancy

```{r}
projection <-
  lc_pred |> 
  select(year, age, lmx_hat) |> 
  mutate(mx = exp(lmx_hat),
         n = rep(1,n()),
         ax = mx_to_ax(mx = mx, age = age, sex = "female")) |> 
  select(-lmx_hat) |> 
  mutate(variant = "lc")

combined_lt <-
  clmx |> 
  mutate(n = 1) |> 
  mutate(variant = "empirical") |> 
  select(year, age, mx, ax, n,variant) |> 
  bind_rows(projection) |> 
  group_by(variant,year) |> 
  reframe(calc_LT(mx = mx, ax = ax, age = age))
```

plot life expectancy:

```{r}
combined_lt |> 
  filter(age == 0) |> 
  ggplot(aes(x = year, y = ex, color = variant)) +
  geom_line() +
  theme_minimal() +
  geom_vline(xintercept = 2020)
```










