---
title: "Session 5 Notes"
author: "Tim Riffe"
date: "2023-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Projection

Pick out some data
```{r}
library(tidyverse)
library(colorspace)
hmd_url <- "https://github.com/timriffe/KOSTAT_Workshop1/raw/master/Data/hmd.csv.gz"
hmd <- read_csv(hmd_url, 
                show_col_types = FALSE)

AUS <- hmd |> 
  filter(country == "Australia",
         sex == "f") |> 
  mutate(lmx = log(mx))
```

## Take a quick look:

```{r}
AUS |> 
  ggplot(mapping = aes(x = year, 
                       y = age, 
                       z = lmx)) +
  geom_contour_filled() +
  coord_equal()
```

# derive alpha

To calculate $\alpha(x)$ just take the mean over time of the log mortality rates in each age. We can do this using `group_by()` and `summarize()`. The line we get, we will use as-is. Notice it's already very smooth, but we risk amplifying any deviations (little bumps and divets) if we project this over time. This gets more serious with $\beta(x)$

```{r}
alpha <- 
  AUS |> 
  group_by(age) |> 
  summarize(alphax = mean(lmx))

# take a look
alpha |> 
  ggplot(mapping = aes(x = age, y = alphax)) +
  geom_line() +
  theme_minimal()
```

# get centered log mortality as matrix
```{r}
cent_l <-
  AUS |> 
  left_join(alpha,by="age") |> 
  mutate(lmx_centered = lmx - alphax) |> 
  select(age, year, lmx_centered) 

cent_l |> 
  ggplot(mapping = aes(x = year, y = age, z = lmx_centered)) +
  geom_contour_filled()+
  coord_equal()

cent_l |> 
  filter(age - age %% 10 == 0) |> 
  ggplot(mapping = aes(x = year, y = lmx_centered, color = age, group = age)) +
  geom_line() +
  scale_color_binned_sequential("Inferno")

```
Turn it into a matrix
```{r}
cenlM <-
  cent_l |> 
  pivot_wider(names_from = year,
              values_from = lmx_centered) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

# perform SVD

```{r}
svd_lM <- svd(cenlM)
str(svd_lM)

U <- svd_lM$u
D <- svd_lM$d
V <- svd_lM$v

# can be recombined:
# (U %*% diag(D) %*% t(V)) - cent_lM

```

# derive beta and kappa

```{r}
u      <- svd_lM$u[, 1]
v      <- svd_lM$v[, 1]
d      <- svd_lM$d[1]

Bx     <- u / sum(u)

# secular change
Kt     <- v * sum(u) * d
```

# put these back in tidy tables:

Recall, a `tibble()` can be created by hand like so:
```{r}
beta <- 
  tibble(betax = Bx,
       age = 0:110)

kappa <- 
  tibble(kappat = Kt,
         year = 1921:2020)
```

# combine to predict:

$$ \widehat{ln(m(x,t))} = \alpha(x)+\beta(x)\cdot\kappa(t)$$

```{r}
fit <-
  cross_join(beta, kappa) |> 
  left_join(alpha, by = "age") |> 
  mutate(lmx_lc = alphax + betax * kappat) |> 
  select(year, age,lmx_lc ) |> 
  left_join(AUS, by = c("age","year")) 

fit |> 
  ggplot(mapping = aes(x = year, y = age, z = lmx_lc)) +
  geom_contour_filled() +
  coord_equal()
```

# examine the residuals

```{r}
fit |> 
  mutate(resid = lmx_lc - lmx) |> 
  ggplot(mapping = aes(x = year, y = age, fill = resid)) +
  geom_tile() +
  scale_fill_continuous_diverging()
```

# zoom on central residuals

Here we see actual patterns that the LC model will miss; We also see patterns of propagation that will not carry on into the future when we forecast.
```{r}
fit |> 
  mutate(resid = lmx_lc - lmx,
         resid = if_else(resid > .7,.7,resid),
         resid = if_else(resid < -.7, -.7, resid)) |> 
  ggplot(mapping = aes(x = year, y = age, fill = resid)) +
  geom_tile() +
  scale_fill_continuous_diverging()
```

# calculate drift and project kappa
```{r}
lK      <- length(Kt)
drift   <- (Kt[lK] - Kt[1]) / (lK - 1)
horizon <- 1:100
Kt.fcst <- Kt[lK] + drift * horizon
```

# recombine to create projected mortality schedule

```{r}
Kappaf <- tibble(year = 2021:2120, kappat = Kt.fcst)

forecast <-
  cross_join(Kappaf, alpha) |> 
  left_join(beta, by = "age") |> 
  mutate(lmx_f = alphax + betax * kappat)
```

# check age patterns
```{r}
forecast |> 
  filter(year %% 10 == 1) |> 
  mutate(mx_f = exp(lmx_f)) |> 
  ggplot(mapping = aes(x = age, y = mx_f, color = year, group = year)) +
  geom_line() +
  scale_y_log10()
```

```{r}
source("https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/lifetable_utils.R")

forecast |> 
  group_by(year) |> 
  mutate(mx = exp(lmx_f),
         n = 1,
         ax = if_else(age == 0, .1, .5),
         qx = calc_qx(mx, ax, n),
         lx = calc_lx(qx),
         dx = calc_dx(lx, qx),
         Lx = calc_Lx(lx, dx, ax, n),
         Tx = calc_Tx(Lx),
         ex = Tx / lx) |> 
  filter(age == 0) |> 
  ggplot(mapping = aes(x = year, y = ex)) +
  geom_line()
```


