---
title: "Intro and configuration of git / github for R"
author: "Tim Riffe"
date: "2022-08-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

In this short session, we aim to get attendees configured to *push* and *pull* between a local RStudio project and a github repository. We will cover initial setup. I will assume that you have installed `git` and have created a github account.

# Configuration

```{r}

#install.packages("usethis")
#install.packages("credentials")

```


## generate personal access token in GitHub (`PAT`)

```{r, eval = FALSE}
library(usethis)
library(credentials)

create_github_token() 
```


## store the `PAT` locally.
```{r, eval = FALSE}
set_github_pat()
```

# Setup

## Clone a project from GitHub 

Click `File | New Project | Version Control | Git ` and then paste the repo url (`.git`) into the url box, and click `create project`.

## Turn an RStudio project into a git repository

This is pretty straightforward, you'll be asked questions in the console. Select NOT to commit all files. Do this deliberately one file at a time later.
```{r, eval = FALSE}
usethis::use_git()
```


## Create a GitHub repository from an RStudio git-controlled project

# General usage advice

## What to commit

## When to commit

## When to go public

# Not covered

There will not be time today to cover topics such as branching, merging, pull requests, or managing conflicts. This is really just a set-up guide, that's all.