---
title: "Session 4 notes"
author: "Tim Riffe"
date: "7/29/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `ggplot2` basics

```{r}
# install.packages("gapminder")
library(gapminder)
library(tidyverse)

gapminder
```

The fundamentals, `ggplot()` is necessary but not sufficient to build the plot. here we declare, at minimum, the coordinate mapping. Executing this opens a blank plot area, which is however aware of the data dimensions.
```{r}
gapminder %>% 
  ggplot(mapping = aes(x = gdpPercap, y = lifeExp))
```

To actually *draw* something (sorry, map something) to the plot, we need to specify a geometric element, using `geom_*`
```{r}
gapminder %>% 
  ggplot(mapping = aes(x = gdpPercap, y = lifeExp)) +
  
  # this declare the geometric mapping
  geom_point()
```

If we instead specify `geom_line()` it renders the observations as a single path, because it doesn't know where each country begins and ends.
```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, 
             y = lifeExp)) +
  #geom_line()
  geom_point(mapping = aes(color = continent), 
             alpha = .5) + 
  scale_x_log10() +
  #scale_y_log10() +
  geom_smooth() # ?geom_smooth
```

Time to map `alpha`, but without legend

```{r}
gapminder %>% 
  ggplot(aes(x = gdpPercap, 
             y = lifeExp)) +
  geom_point(mapping = aes(color = continent,
                           alpha = year)) +
  scale_x_log10() +
  # turn off alpha legend
  guides(alpha = "none")
```

Add some filters to look at just one subset:

```{r}
gapminder %>% 
  # filter down to large countries in Africa
  filter(continent == "Africa",
         pop > 1e7) %>% 
  
  # start the plot, declaring only coordinates
  # so that we can have aesthetic control over each geom
  ggplot(aes(x = gdpPercap,
             y = lifeExp)) +
  
  # x and y map to points, whose color *depends on* the country
  # and transparency on year, and size on population
  geom_point(mapping = aes(color = country,
                           alpha = year,
                           size = pop)) +
  
  # same mapping for lines, except we are explicit that 
  # countries are distinct groups, so each line should
  # be separate
  geom_line(mapping = aes(group = country,
                          color = country,
                          alpha = year)) +
  
  # turn off color and transparency legends
  guides(alpha = "none",
         color = "none") +
  
  # log the x scale
  scale_x_log10() +
  
  # add overall smoother with explicit aesthetic settings
  geom_smooth(color = "red",
              fill = "red",
              alpha = .2)
```

I'm removing the smoother because it's distracting in a bad way: it doesn't summarize the within-country trend the way we'd like.

```{r}
# install.packages("scales")
library(scales)
library(ggthemes)
gapminder %>% 
  filter(pop > 1e7,
         continent != "Oceania") %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(mapping = aes(color = country,
                           size = pop,
                           alpha = year))+
  geom_line(mapping = aes(color = country,
                          alpha = year,
                          group = country))+
  scale_x_log10() + 
  scale_size_continuous(labels = number_format(scale = 1 / 1e6)) +
  guides(color = "none",
         alpha = "none") +
  facet_wrap(~continent) +
  labs(size = "Population\n(millions)",
       x = "GDP per capita",
       y = "Life expectancy at birth",
       title = "Trends in GDP per capita and life expectancy",
       subtitle = "Data collated by gapminder") +
  theme_fivethirtyeight()
  
# install.packages("ggthemes")
```

# Visualize our lifetables from Wednesday

```{r}
library(readr)
path <- "https://raw.githubusercontent.com/timriffe/KOSTAT_Workshop1/master/Data/AFR_LT.csv"
AFR_LT <- read_csv(path)
```

Examine the data, plot all $M_x$ curves in a way that is legible

```{r}
AFR_LT %>% head()

AFR_LT %>% 
  filter(Year == max(Year),
         Sex != "t") %>% 
 ggplot(mapping = aes(x = Age, y = nMx, group = interaction(Country, Sex))) +
 geom_line(aes(color = Sex), alpha = .3) +
 scale_y_log10()
```

Make a sex ratios plot, this requires pre-processing

```{r}
AFR_LT %>% 
  # for sex ratios throw out total,
  # keep only most recent year
  filter(Year == max(Year),
         Sex != "t") %>% 
  
  # select down to only-needed columns before pivot wider,
  # just because I was anticipating a very wide dataset otherwise
  # or at least not the anticipated dimensions
  select(Country, Age, Sex, nMx) %>% 
  
  # move males and females side by side
  pivot_wider(names_from = Sex, values_from = nMx) %>% 
  
  # calculate the measure
  mutate(SR = m / f) %>% 
  
  # here begins the plot
  ggplot(aes(x = Age, y = SR, group = Country)) + 
  geom_line() +
  scale_y_log10()
```

# Population pyramids using `geom_bar()`

Our strategy for making a population pyramid will look a lot like the same steps you'd take anywhere else. Namely, to get males on the left, just make them negative.

```{r}
countries <- AFR_LT$Country %>% unique()
# sample(countries, 1)
AFR_LT %>% 
  filter(Country == "Nigeria",
         Year == 2019,
         Sex != "t") %>% 
  mutate(SRB = 1.05,
         PF = 1 / (1 + SRB),
         `Stationary Population` = ifelse(Sex == "m", 
                                          nLx * (1 - PF),
                                          nLx * PF),
         `Stationary Structure` = 100 * `Stationary Population` / sum(`Stationary Population`),
         `Stationary Structure` = ifelse(Sex == "m",
                                         -`Stationary Structure`,
                                         `Stationary Structure`) / 5,
         Age = Age - Age %% 5
         ) %>%   
  group_by(Sex,Age) %>% 
  summarize(`Stationary Structure` = sum(`Stationary Structure`),
            .groups = "drop") %>% 
  
  ggplot(aes(x = Age,                                                                                           y = `Stationary Structure`,
             fill = Sex,
             color = Sex)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = seq(-.7, .7, by = .2),
                     labels = paste0(c("0.7","0.5","0.3","0.1","0.1","0.3","0.5","0.7"),"%")) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Stationary population structure",
       subitle = "Nigeria, 2019",
       caption = "Data: based on data from GHO")
```












































